---
title: "Feature selection"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/03-feature-selection.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r}
library(BiocSingular)
library(org.Mm.eg.db)
library(eulerr)
library(pheatmap)
library(scales)
library(scater)
library(scran)
library(msigdbr)
library(EnsDb.Mmusculus.v79)
```

Read experiment data:

```{r}
sce <- readRDS("output/02-normalization.rds")
```

## Blacklist

Create a blacklist of genes to ignore for feature selection:

```{r}
rowData(sce)$blacklist <- FALSE
```

Append Xist gene to blacklist:

```{r}
ids <- keys(EnsDb.Mmusculus.v79, keytype = "GENEID", filter = GeneNameFilter("Xist"))

idx <- which(rowData(sce)$gene_id %in% ids)

rowData(sce)$blacklist[idx] <- TRUE
```

Append Y chromosome genes to blacklist:

```{r}
ids <- keys(EnsDb.Mmusculus.v79, keytype = "GENEID", filter = SeqNameFilter("Y"))

idx <- which(rowData(sce)$gene_id %in% ids)

rowData(sce)$blacklist[idx] <- TRUE
```

Append MT chromosome genes to blacklist:

```{r}
ids <- keys(EnsDb.Mmusculus.v79, keytype = "GENEID", filter = SeqNameFilter("MT"))

idx <- which(rowData(sce)$gene_id %in% ids)

rowData(sce)$blacklist[idx] <- TRUE
```

Append cell-cycle genes to blacklist:

```{r}
org <- select(org.Mm.eg.db, keys = "GO:0007049", keytype = "GOALL", column = "ENSEMBL")

idx <- which(rowData(sce)$gene_id %in% org$ENSEMBL)

rowData(sce)$blacklist[idx] <- TRUE
```

Append ribosomal genes to blacklist:

```{r}
sig <- msigdbr(species = "Mus musculus", category = "C2")

sig <- subset(sig, gs_name == "KEGG_RIBOSOME")

org <- select(org.Mm.eg.db, keys = sig$gene_symbol, keytype = "SYMBOL", column = "ENSEMBL")

idx <- which(rowData(sce)$gene_id %in% org$ENSEMBL)

rowData(sce)$blacklist[idx] <- TRUE
```

## Measuring variation {.tabset}

### modelGeneVar

Model the per-gene variance

```{r}
dec <- modelGeneVar(sce, block = sce$stage)

ids <- colnames(dec$per.block)

plt <- lapply(ids, function(i) {
  
  dat <- as.data.frame(dec$per.block[[i]])
  
  ggplot(dat, aes(mean, total)) + 
    geom_point(colour = "grey") + 
    geom_line(aes(y = tech)) + 
    labs(title = i, x = "Mean of log-expression", y = "Variance of log-expression") + 
    theme_minimal()

})

wrap_plots(plt)

metadata(sce)$modelGeneVar <- dec
```

### modelGeneCV2

Model the per-gene CV2:

```{r}
dec <- modelGeneCV2(sce, block = sce$stage)

ids <- colnames(dec$per.block)

plt <- lapply(ids, function(i) {
  
  dat <- as.data.frame(dec$per.block[[i]])
  
  ggplot(dat, aes(mean, total)) + 
    geom_point(colour = "grey") + 
    geom_line(aes(y = trend)) + 
    scale_x_log10() + 
    scale_y_log10() + 
    labs(title = i, x = "Mean of log-expression", y = "Variance of log-expression") + 
    theme_minimal()

})

wrap_plots(plt)

metadata(sce)$modelGeneCV2 <- dec
```

### modelGeneVarWithSpikes

Model the per-gene variance with spike-ins:

```{r eval = FALSE}
dec <- modelGeneVarWithSpikes(sce, spikes = "spike", block = sce$stage)

ids <- colnames(dec$per.block)

plt <- lapply(ids, function(i) {
  
  dat <- dec$per.block[[i]]
  
  alt <- metadata(dec)
  
  df1 <- as.data.frame(dec)
  
  df2 <- data.frame(mean = alt$mean, total = alt$var)
  
  ggplot(df1, aes(mean, total)) + 
    geom_point(colour = "grey") + 
    geom_line(aes(mean, tech)) + 
    geom_point(data = df2, aes(mean, total), colour = "red") + 
    labs(title = i, x = "Mean of log-expression", y = "Variance of log-expression") + 
    theme_minimal()

})

metadata(sce)$modelGeneVarWithSpikes <- dec
```

### modelGeneCV2WithSpikes

Model the per-gene CV2 with spike-ins:

```{r eval = FALSE}
dec <- modelGeneCV2WithSpikes(sce, spikes = "spike", block = sce$stage)

ids <- colnames(dec$per.block)

plt <- lapply(ids, function(i) {
  
  dat <- dec$per.block[[i]]
  
  alt <- metadata(dat)
  
  df1 <- as.data.frame(dat)
  
  df2 <- data.frame(mean = alt$mean, cv2 = alt$cv2)
  
  ggplot(df1, aes(mean, total)) + 
    geom_point(colour = "grey") + 
    geom_line(aes(mean, trend)) + 
    geom_point(data = df2, aes(mean, cv2), colour = "red") + 
    scale_x_log10() + 
    scale_y_log10() + 
    labs(title = i, x = "Mean of log-expression", y = "Variance of log-expression") + 
    theme_minimal()
  
})

wrap_plots(plt)

metadata(sce)$modelGeneCV2WithSpikes <- dec
```

### modelGeneVarByPoisson

Model the per-gene variance with Poisson noise:

```{r}
dec <- modelGeneVarByPoisson(sce, block = sce$stage)

ids <- colnames(dec$per.block)

plt <- lapply(ids, function(i) {
  
  dat <- as.data.frame(dec$per.block[[i]])
  
  ggplot(dat, aes(mean, total)) + 
    geom_point(colour = "grey") + 
    geom_line(aes(y = tech)) + 
    labs(title = i, x = "Mean of log-expression", y = "Variance of log-expression") + 
    theme_minimal()

})

wrap_plots(plt)

metadata(sce)$modelGeneVarByPoisson <- dec
```

## Selecting genes {.tabset}

### modelGeneVar

Identify highly variable genes:

```{r}
dec <- metadata(sce)$modelGeneVar

hvg <- getTopHVGs(dec, var.field = "bio", var.threshold = 0, fdr.field = "FDR", fdr.threshold = 0.05)

del <- rownames(sce)[rowData(sce)$blacklist == TRUE]

hvg <- setdiff(hvg, del)

sel <- rownames(dec) %in% hvg

metadata(sce)$modelGeneVar$hvg <- sel
```

Plot expression of highly variable genes:

```{r}
plotHeatmap(sce, hvg, center = TRUE, symmetric = TRUE, show_rownames = FALSE, show_colnames = FALSE)
```

### modelGeneCV2

Identify highly variable genes:

```{r}
dec <- metadata(sce)$modelGeneCV2

hvg <- getTopHVGs(dec, var.field = "ratio", var.threshold = 1, fdr.field = "FDR", fdr.threshold = 0.05)

del <- rownames(sce)[rowData(sce)$blacklist == TRUE]

hvg <- setdiff(hvg, del)

sel <- rownames(dec) %in% hvg

metadata(sce)$modelGeneCV2$hvg <- sel
```

Plot expression of highly variable genes:

```{r}
plotHeatmap(sce, hvg, center = TRUE, symmetric = TRUE, show_rownames = FALSE, show_colnames = FALSE)
```

### modelGeneVarWithSpikes

Identify highly variable genes:

```{r eval = FALSE}
dec <- metadata(sce)$modelGeneVarWithSpikes

hvg <- getTopHVGs(dec, var.field = "bio", var.threshold = 0, fdr.field = "FDR", fdr.threshold = 0.05)

del <- rownames(sce)[rowData(sce)$blacklist == TRUE]

hvg <- setdiff(hvg, del)

sel <- rownames(dec) %in% hvg

metadata(sce)$modelGeneVarWithSpikes$hvg <- sel
```

Plot expression of highly variable genes:

```{r eval = FALSE}
plotHeatmap(sce, hvg, center = TRUE, symmetric = TRUE, show_rownames = FALSE, show_colnames = FALSE)
```

### modelGeneCV2WithSpikes

Identify highly variable genes:

```{r eval = FALSE}
dec <- metadata(sce)$modelGeneCV2WithSpikes

hvg <- getTopHVGs(dec, var.field = "ratio", var.threshold = 1, fdr.field = "FDR", fdr.threshold = 0.05)

del <- rownames(sce)[rowData(sce)$blacklist == TRUE]

hvg <- setdiff(hvg, del)

sel <- rownames(dec) %in% hvg

metadata(sce)$modelGeneCV2WithSpikes$hvg <- sel
```

Plot expression of highly variable genes:

```{r eval = FALSE}
plotHeatmap(sce, hvg, center = TRUE, symmetric = TRUE, show_rownames = FALSE, show_colnames = FALSE)
```

### modelGeneVarByPoisson

Identify highly variable genes:

```{r}
dec <- metadata(sce)$modelGeneVarByPoisson

hvg <- getTopHVGs(dec, var.field = "bio", var.threshold = 0, fdr.field = "FDR", fdr.threshold = 0.05)

del <- rownames(sce)[rowData(sce)$blacklist == TRUE]

hvg <- setdiff(hvg, del)

sel <- rownames(dec) %in% hvg

metadata(sce)$modelGeneVarByPoisson$hvg <- sel
```

Plot expression of highly variable genes:

```{r}
plotHeatmap(sce, hvg, center = TRUE, symmetric = TRUE, show_rownames = FALSE, show_colnames = FALSE)
```

## Comparison

### Euler

Compare number of features selected using an Euler diagram:

```{r}
sel <- cbind(
  "Features" = TRUE,
  "Var" = metadata(sce)$modelGeneVar$hvg,
  "CV2" = metadata(sce)$modelGeneCV2$hvg,
  # "VarWithSpikes" = metadata(sce)$modelGeneVarWithSpikes$hvg,
  # "CV2WithSpikes" = metadata(sce)$modelGeneCV2WithSpikes$hvg,
  "VarByPoisson" = metadata(sce)$modelGeneVarByPoisson$hvg
)

fit <- euler(sel)

pal <- RColorBrewer::brewer.pal(n = 5, name = "Pastel1")

lab <- comma(fit$original.values)

plot(fit, fills = pal, quantities = lab)
```

### Jaccard

Calculate Jaccard index:

```{r}
sel <- rbind(
  "Var" = metadata(sce)$modelGeneVar$hvg,
  "CV2" = metadata(sce)$modelGeneCV2$hvg,
  # "VarWithSpikes" = metadata(sce)$modelGeneVarWithSpikes$hvg,
  # "CV2WithSpikes" = metadata(sce)$modelGeneCV2WithSpikes$hvg,
  "VarByPoisson" = metadata(sce)$modelGeneVarByPoisson$hvg
)

dst <- proxy::dist(sel, method = "Jaccard", diag = TRUE, upper = TRUE)

mat <- as.matrix(dst)

pal <- RColorBrewer::brewer.pal(n = 5, name = "Greens")

col <- colorRampPalette(pal)(10)

brk <- seq(0, 1, length.out = 11)

pheatmap(mat, color = col, breaks = brk)
```

### Dimensionality reduction {.tabset}

```{r}
dec <- c(
  "modelGeneVar",
  "modelGeneCV2",
  # "modelGeneVarWithSpikes",
  # "modelGeneCV2WithSpikes",
  "modelGeneVarByPoisson"
)
```

#### PCA

```{r, fig.height = 11, fig.width = 10}
plt <- lapply(dec, function(x) {
  
  set.seed(0000111001)
  
  hvg <- metadata(sce)[[x]]$hvg
  
  sce <- runPCA(sce, subset_row = hvg, BSPARAM = ExactParam())

  plotPCA(sce, colour_by = "sum") + ggtitle(x)

})

patchwork::wrap_plots(plt, ncol = 2)
```

#### TSNE

```{r, fig.height = 11, fig.width = 10}
plt <- lapply(dec, function(x) {
  
  set.seed(0100011011)
  
  hvg <- metadata(sce)[[x]]$hvg
  
  sce <- runPCA(sce, subset_row = hvg, BSPARAM = ExactParam())
  
  sce <- runTSNE(sce, dimred = "PCA")

  plotTSNE(sce, colour_by = "sum") + ggtitle(x)

})

patchwork::wrap_plots(plt, ncol = 2)
```

#### UMAP

```{r, fig.height = 11, fig.width = 10}
plt <- lapply(dec, function(x) {
  
  set.seed(0101001100)
  
  hvg <- metadata(sce)[[x]]$hvg
  
  sce <- runPCA(sce, subset_row = hvg, BSPARAM = ExactParam())
  
  sce <- runUMAP(sce, dimred = "PCA")

  plotUMAP(sce, colour_by = "sum") + ggtitle(x)

})

patchwork::wrap_plots(plt, ncol = 2)
```

## Selection

Select highly variable genes and store in row metadata:

```{r}
dec <- metadata(sce)$modelGeneVarWithSpikes

dec <- subset(dec, hvg == TRUE)

sel <- rownames(sce) %in% rownames(dec)

rowData(sce)$selected <- sel
```

## Summary

### Output

Save experiment object to disk:

```{r}
saveRDS(sce, "output/03-feature-selection.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
