---
title: "Data integration"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Setup chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/04-data-integration.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load Bioconductor packages:

```{r}
library(batchelor)
library(EnsDb.Mmusculus.v79)
library(scater)
library(scran)
library(scuttle)
```

Load CRAN packages:

```{r}
library(ggforce)
library(pheatmap)
```

Read experiment data:

```{r}
sce <- readRDS("output/03-feature-selection.rds")
```

## Integration

Retrieve highly variable genes:

```{r}
hvg <- rownames(sce)[rowData(sce)$selected]
```

Perform scaling normalization:

```{r}
sce <- multiBatchNorm(sce, batch = sce$stage)
```

Correct experiments using nearest neighbours:

```{r}
set.seed(0111110100)

mnn <- correctExperiments(
  sce,
  batch = sce$stage,
  subset.row = hvg,
  correct.all = TRUE,
  PARAM = FastMnnParam(merge.order = c("E9.5", "E8.5"), get.variance = TRUE)
)
```

## Diagnostics

Check proportion of lost variance:

```{r}
var <- metadata(mnn)$merge.info$lost.var

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Reds"))(100)

brk <- seq(0, max(var, na.rm = TRUE), length.out = 101)

pheatmap(var, color = col, display_numbers = TRUE, number_format = "%.2f", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Cluster on the corrected PCA to obtain a partitioning of the cells:

```{r}
snn <- buildSNNGraph(mnn, type = "jaccard", use.dimred = "corrected")

com <- igraph::cluster_louvain(snn)

com$membership <- as.factor(com$membership)
```

Tabulate number of cells from multiple batches within each cluster:

```{r}
tab <- table(Cluster = com$membership, Batch = mnn$batch)

col <- colorRampPalette(RColorBrewer::brewer.pal(n = 5, name = "Blues"))(100)

brk <- seq(0, max(tab), length.out = 101)

pheatmap(tab, color = col, display_numbers = TRUE, number_format = "%d", cluster_cols = FALSE, cluster_rows = FALSE, angle_col = 0)
```

Visualize the corrected PCA:

```{r}
ggcells(mnn, aes(corrected.1, corrected.2, colour = com$membership)) + 
  geom_point(size = 1) + 
  scale_colour_brewer(name = "Stage", palette = "Set1") + 
  labs(x = "PCA 1", y = "PCA 2") + 
  facet_wrap(~ batch) + 
  theme_no_axes() + 
  coord_fixed()
```

Perform TSNE on the corrected PCA matrix:

```{r}
set.seed(1111100110)

mnn <- runTSNE(mnn, dimred = "corrected")
```

Visualize the corrected PCA using a TSNE plot:

```{r}
ggcells(mnn, aes(TSNE.1, TSNE.2, colour = com$membership)) +
  geom_point(size = 1) + 
  scale_colour_brewer(name = "Stage", palette = "Set1") + 
  labs(x = "TSNE 1", y = "TSNE 2") +
  facet_wrap(~ batch) +
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

Perform UMAP on the corrected PCA matrix:

```{r}
set.seed(1001010111)

mnn <- runUMAP(mnn, dimred = "corrected")
```

Visualize the corrected PCA using a UMAP plot:

```{r}
ggcells(mnn, aes(UMAP.1, UMAP.2, colour = com$membership)) +
  geom_point(size = 1) +
  scale_colour_brewer(name = "Stage", palette = "Set1") + 
  labs(x = "UMAP 1", y = "UMAP 2") +
  facet_wrap(~ batch) +
  theme_no_axes() + 
  theme(aspect.ratio = 1)
```

## Summary

### Output

Save experiment object to disk:

```{r}
saveRDS(mnn, file = "output/04-data-integration.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
