---
title: "Doublet detection"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/06-doublet-detection.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load Bioconductor packages:

```{r}
library(scater)
library(scran)
library(scuttle)
```

Read experiment data:

```{r}
sce <- readRDS("output/05-clustering.rds")
```

## Detection

### By cluster {.tabset}

Check there are at least three clusters to detect doublet clusters:

```{r}
num <- nlevels(sce$cluster)

run <- ifelse(num < 3, FALSE, TRUE)
```

Identify potential clusters of doublet cells based on intermediate expression profiles:

```{r eval = run}
dbl <- doubletCluster(sce, sce$cluster)
```

Annotate clusters that have unusually low number of unique genes:

```{r eval = run}
out <- isOutlier(dbl$N + 1, type = "lower", log = TRUE) # add pseudocount for log transform

ids <- rownames(dbl)[out]

sce$doublet_cluster <- sce$cluster %in% ids
```

Check for clusters that have unusually low number of unique genes:

```{r eval = run}
run <- any(sce$doublet_cluster)
```

Find significant marker genes between clusters:

```{r eval = run}
res <- findMarkers(sce, sce$cluster, direction = "up")

sig <- lapply(res, subset, FDR < 0.1)

sig <- Filter(nrow, sig)
```

Inspect the expression of marker genes across clusters:

```{r eval = run, results = "asis"}
for (n in names(sig)) {
  
  dat <- sig[[n]]
  
  dat <- subset(dat, Top <= 10)
  
  ids <- rownames(dat)
  
  cat("#### Cluster", n, "\n")

  plotHeatmap(
    x = sce, 
    features = ids, 
    center = TRUE,
    symmetric = TRUE, 
    color_columns_by = "stage",
    order_columns_by = "cluster"
  )

  cat("\n\n")

}
```

### By cell {.tabset}

Identify potential doublet cells based on simulations of putative doublet expression profiles:

```{r}
set.seed(0110101011)

dim <- reducedDim(sce)

num <- ncol(dim)

dbl <- doubletCells(sce, d = num)

sce$doublet_score <- log10(dbl + 1)
```

Inspect the distribution of doublet scores on reduced dimensions:

#### PCA

```{r}
plotReducedDim(sce, dimred = "PCA", colour_by = "doublet_score", shape_by = "stage")
```

#### TSNE

```{r}
plotTSNE(sce, colour_by = "doublet_score", shape_by = "stage")
```

#### UMAP

```{r}
plotUMAP(sce, colour_by = "doublet_score", shape_by = "stage")
```

## Summary

### Output

Save experiment object to disk:

```{r}
saveRDS(sce, file = "output/06-doublet-detection.rds" )
```

### Session

Print session information:

```{r}
sessionInfo()
```
