---
title: "Cell cycle assignment"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Define chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/08-cell-cycle.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load Bioconductor packages:

```{r}
library(cyclone)
library(scran)
library(scater)
library(scuttle)
library(org.Mm.eg.db)
```

Import experiment data:

```{r}
sce <- readRDS("output/07-doublet-detection.rds")
```

## Expression

Inspect the relative expression of cyclins across the clusters:

```{r}
idx <- grep("^Ccn[abde][0-9]$", rowData(sce)$gene_name)

ids <- rownames(sce)[idx]

plotHeatmap(sce, features = sort(ids), colour_columns_by = "stage", order_columns_by = "cluster", cluster_rows = FALSE)
```

Inspect the relative expression of cell-cycle related genes across the clusters:

```{r}
org <- select(org.Mm.eg.db, keys = "GO:0007049", keytype = "GOALL", column = "ENSEMBL")

idx <- which(rowData(sce)$gene_id %in% org$ENSEMBL)

ids <- rownames(sce)[idx]

plotHeatmap(sce, features = ids, colour_columns_by = "stage", order_columns_by = "cluster", center = TRUE, symmetric = TRUE, show_rownames = FALSE)
```

## Cyclone {.tabset}

Perform cell cycle phase classification:

```{r}
con <- system.file("exdata", "mouse_cycle_markers.rds", package = "scran")

ids <- readRDS(con)

fit <- cyclone(sce, pairs = ids, gene.names = rowData(sce)$gene_id)
```

Assign cell cycle phase:

```{r}
sce$phase <- fit$phases
```

Plot cell cycle phase scores:

```{r}
ggplot(fit$scores, aes(G1, G2M, colour = fit$phases)) + 
  geom_point() + 
  scale_colour_brewer(name = "Phase", palette = "Set1") + 
  theme_bw()
```

Plot cell cycle phase by cluster:

```{r}
dat <- makePerCellDF(sce, use.coldata = c("cluster", "phase"))

ggplot(dat, aes(cluster)) + 
  geom_bar(aes(fill = phase)) + 
  scale_fill_brewer(name = "Phase", palette = "Set1") + 
  labs(x = "Cluster", y = "Count") + 
  theme_bw()
```

Plot cell cycle phase by embedding:

### PCA

```{r}
plotReducedDim(sce, dimred = "corrected", colour_by = "phase", shape_by = "stage", text_by = "cluster")
```

### TSNE

```{r}
plotTSNE(sce, colour_by = "phase", shape_by = "stage", text_by = "cluster")
```

### UMAP

```{r}
plotUMAP(sce, colour_by = "phase", shape_by = "stage", text_by = "cluster")
```

## Summary

### Output

Save experiment object to disk:

```{r}
saveRDS(sce, file = "output/08-cell-cycle.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
