---
title: "Cell type annotation"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Define chunk options:

```{r}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.lazy = FALSE,
  cache.path = "cache/10-cell-annotation.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align = "center",
  out.width = '100%'
)
```

Load Bioconductor packages:

```{r}
library(scater)
library(scran)
library(scuttle)
library(SingleR)
```

Load CRAN packages:

```{r}
library(patchwork)
library(pheatmap)
```

Source custom functions:

```{r}
source("code/cleanEmbryoAtlasData.R")
```

Import experiment data:

```{r}
sce <- readRDS("output/09-marker-detection.rds")
```

## Annotation

### Reference data

Load the mouse gastrulation data package:

```{r}
library(MouseGastrulationData)
```

Subset atlas sample metadata by embryonic stage:

```{r}
AtlasSampleMetadata <- subset(AtlasSampleMetadata, stage == "E8.5")
```

Retrieve processed samples from embryo atlas data:

```{r}
ref <- EmbryoAtlasData(type = "processed", samples = AtlasSampleMetadata$sample)
```

Clean processed samples from embryo atlas data:

```{r}
ref <- cleanEmbryoAtlasData(ref)
```

Identify marker genes for each cell type whilst blocking on sample:

```{r}
res <- pairwiseTTests(logcounts(ref), groups = ref$celltype, block = ref$sample)

sig <- getTopMarkers(res$statistics, res$pairs)
```

Aggregate references within each celltype and sample combination:

```{r}
lab <- paste(ref$celltype, ref$sample, sep = "-")

ref <- aggregateReference(ref, labels = lab, BSPARAM = BiocSingular::ExactParam())
```

Remove sample label from aggregated reference names:

```{r}
lab <- sapply(strsplit(colnames(ref), "-"), .subset, 1)
```

Return the best annotation for each cell in the test dataset:

```{r}
ids <- rownames(sce) # save original feature names

rownames(sce) <- rowData(sce)$gene_id # make feature names compatible

fit <- SingleR(sce, ref, labels = lab, genes = sig)

rownames(sce) <- ids # return original feature names
```

Store SingleR assignments in experiment data:

```{r}
sce$celltype_singler <- fit$labels
```

Create a heatmap of the average SingleR assignment scores:

```{r}
plotScoreHeatmap(fit, clusters = sce$cluster, fontsize = 8)
```

Plot score distributions of labels:

```{r}
plotScoreDistribution(fit, show = "delta.med", ncol = 3, show.nmads = 3)
```

## Selection

Choose cell type annotation method for downstream analysis:

```{r}
sce$celltype <- sce$celltype_singler
```

## Summary

### Output

Save experiment object to disk:

```{r}
saveRDS(sce, file = "output/10-cell-annotation.rds")
```

### Session

Print session information:

```{r}
sessionInfo()
```
