---
title: "Interactive data exploration"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set chunk options:

```{r}
knitr::opts_chunk$set(eval = TRUE)
```

Load CRAN packages:

```{r}
library(rsconnect)
```

Include Bioconductor package repositories:

```{r}
options(repos = BiocManager::repositories())
```

Automatically answer yes to any user prompt:

```{r}
options(needs.promptUser = FALSE)
```

Delete existing shiny directory:

```{r}
unlink("shiny", recursive = TRUE)
```

Create shiny directory for interactive data exploration:

```{r}
dir.create("shiny", showWarnings = FALSE, mode = "0755")
```

## Application {.tabset}

Copy data to iSEE directory:

```{r}
file.copy("output/10-cell-annotation.rds", "shiny/data.rds", overwrite = TRUE)
```

Write code to iSEE application file:

```{cat, engine.opts = list(file = "shiny/app.R"), class.source = "r"}
# Load required packages
library(iSEE)
library(SingleCellExperiment)

# Load experiment object
sce <- readRDS("data.rds")

# Clear metadata list
metadata(sce) <- list()

# ReOrder assays
assays(sce) <- assays(sce)[c("counts", "logcounts", "reconstructed")]

# Rescale reconstructed assay
assay(sce, "rescaled") <- scales::rescale(x = as.matrix(assay(sce, "reconstructed")), to = c(0, 100))

# Redefine Brachyury feature name
ind <- which(rowData(sce)$gene_id == "ENSMUSG00000062327")
rowData(sce)$gene_name[ind] <- "Bra"

# Make feature names unique
rownames(sce) <- scuttle::uniquifyFeatureNames(rowData(sce)$gene_id, rowData(sce)$gene_name)

# Clear rowData variables
rowData(sce)$rotation <- NULL

# Define population markers
sce$Bra <- counts(sce)["Bra", ] > 0
sce$Sox2 <- counts(sce)["Sox2", ] > 0
sce$Meox1 <- counts(sce)["Meox1", ] > 0
sce$Msgn1 <- counts(sce)["Msgn1", ] > 0
sce$Tbx6 <- counts(sce)["Tbx6", ] > 0
sce$NMP <- sce$Bra & sce$Sox2
sce$MPC <- sce$Bra & sce$ Msgn1 & sce$Tbx6
sce$PSM <- !sce$Bra & sce$Msgn1 & sce$Tbx6

# Rename and reorder reduced dimensions
reducedDimNames(sce) <- c("MNN", "PCA", "TSNE", "UMAP")
reducedDims(sce) <- reducedDims(sce)[c("PCA", "MNN", "TSNE", "UMAP")]

# Define column data colours
celltype_colours <- function(n) {
  c(
    "Allantois" = "#532C8A",
    "Anterior Primitive Streak" = "#C19F70",
    "Blood progenitors 1" = "#D9DECF",
    "Blood progenitors 2" = "#C9A997",
    "Cardiomyocytes" = "#B51D8D",
    "Caudal epiblast" = "#9E6762",
    "Caudal Mesoderm" = "#3F84AA",
    "Caudal neurectoderm" = "#354E23",
    "Def. endoderm" = "#F397C0",
    "Endothelium" = "#FF891C",
    "Epiblast" = "#635547",
    "Erythroid1" = "#C72228",
    "Erythroid2" = "#F79083",
    "Erythroid3" = "#EF4E22",
    "ExE ectoderm" = "#989898",
    "ExE endoderm" = "#7F6874",
    "ExE mesoderm" = "#8870AD",
    "Forebrain/Midbrain/Hindbrain" = "#647A4F",
    "Gut" = "#EF5A9D",
    "Haematoendothelial progenitors" = "#FBBE92",
    "Intermediate mesoderm" = "#139992",
    "Mesenchyme" = "#CC7818",
    "Mixed mesoderm" = "#DFCDE4",
    "Nascent mesoderm" = "#C594BF",
    "Neural crest" = "#C3C388",
    "NMP" = "#8EC792",
    "Notochord" = "#0F4A9C",
    "Paraxial mesoderm" = "#8DB5CE",
    "Parietal endoderm" = "#1A1A1A",
    "PGC" = "#FACB12",
    "Pharyngeal mesoderm" = "#C9EBFB",
    "Primitive Streak" = "#DABE99",
    "Rostral neurectoderm" = "#65A83E",
    "Somitic mesoderm" = "#005579",
    "Spinal cord" = "#CDE088",
    "Surface ectoderm" = "#F7F79E",
    "Visceral endoderm" = "#F6BFCB"
  )
}

boolean_colours <- function(n) {
  c("TRUE" = "#E15759", "FALSE" = "#BAB0AC")
}

# Set iSEE options
iSEEOptions$set(downsample = TRUE)
iSEEOptions$set(color.maxlevels = 37)

# Define iSEE colours
ecm <- ExperimentColorMap(
  colData = list(
    celltype = celltype_colours,
    Bra = boolean_colours, 
    Sox2 = boolean_colours, 
    Meox1 = boolean_colours, 
    Msgn1 = boolean_colours, 
    Tbx6 = boolean_colours, 
    NMP = boolean_colours, 
    MPC = boolean_colours, 
    PSM = boolean_colours, 
  )
)

# Create iSEE object
iSEE(sce, appTitle = "e-NMPs CLE (E8.5/E9.5) [MNN]", colormap = ecm)
```

Deploy the iSEE application to the shiny server:

```{r}
deployApp(appDir = "shiny", appName = "GOUTI_2017_NMP_MNN", appTitle = "e-NMPs CLE (E8.5/E9.5) [MNN]", launch.browser = FALSE, forceUpdate = TRUE)
```

Allocate more memory to the iSEE application:

```{r}
configureApp(appName = "GOUTI_2017_NMP_MNN", appDir = "shiny", size = "xlarge")
```

## Summary

### Session

Print session information:

```{r}
sessionInfo()
```
